### 2016-09-06 capability-machine.tex
Some notes on OS-less stack and return pointer handling.

>>>>

\section{Other examples and applications}
\label{sec:other_apps}
This section contains some ideas about other examples and applications than the
ticket dispenser example.

\subsection{Stack and return pointer handling without OS involvement using local
  capabilities}
The idea of this example would be to work out and prove a calling convention
that enforces well-bracketed control flow and encapsulation of local variables
using CHERI's local capabilities.

When one function invokes another function, the essential idea is that:
\begin{itemize}
\item Stack pointer is passed as a local and store-local capability.
\item Return pointer is passed as a local capability.
\end{itemize}

Since local pointers cannot leave the registers except into regions for which a
store-local capability is available, this basic idea seems to enforce a number
of useful properties: well-bracketedness of control flow and encapsulation of
private state stored on the stack. On the other hand, it also seems to validate
the standard C treatment of the stack: the stack can be reused after a function
returns, even between distrusting parties. However, safety/security of this
design is very non-trivial and seems to rely on some non-trivial reasoning:

\paragraph{Only stack is store-local?}
A critical assumption is that adversary code has no way to \emph{store} local
capabilities except on the stack. The reason that it is fine to store local
capabilities on the stack is that the adversary only has a \emph{local}
capability to the stack and cannot usefully store that capability anywhere.
However, this means that we need to rely on the runtime system of our
programming language to be careful when handing out store-local capabilities:
only the libc startup code should initialize the stack as store-local and malloc
should \emph{not} produce them. This basically means that the libc
initialization code (or whatever component produces the initial stack pointer)
is part of our TCB.

\paragraph{Requirement for clearing the stack}
Imagine the following trusted C function:

\begin{verbatim}
void myfunction(){
  advfunction1();
  advfunction2();
}
\end{verbatim}

where advfunction1() and advfunction2() are adversary functions. In the standard
C treatment of the stack, advfunction2() would get the same stack pointer as
advfunction1(). This is supposed to be safe since advfunction1() must have
released all capabilities to the stack after its execution. But what if we
require that the two functions have no way of communicating with each other?
Concretely, advfunction1() has access to some secrets that must not be leaked to
advfunction2(). How can we prevent advfunction1() from storing the secret
somewhere on the stack and relying on advfunction2() from receiving the same
stack pointer where it can read the secret? The most obvious solution seems to
be that we should fully clear the stack (overwrite it with zeros) after the
return of any adversary function, but this could cause an important overhead.
Perhaps the processor should accomodate this with a special instruction that can
zero the entire array that a capability points to?

\paragraph{What do return pointers look like?}
An important question is what return pointers look like? Since we want to
protect the caller from the callee, it's important that the return pointer is
opaque, i.e. an entry pointer. The entry pointer will point to a closure that
contains the next instruction to execute, as well as the previous stack pointer.
But since stack pointers are local, this means that the return pointer closure
should be stored in a region of memory for which we have store-local permission,
i.e. on the stack. This means we need the following in our calling convention:
before invoking a function, we push the stack pointer and the instruction
pointer after invocation on the stack, we construct a return pointer by
copying the stack pointer, limiting it to these two entries and making it an
entry pointer.  Then we shrink the stack pointer to the unused part of the stack
and jump.

\paragraph{Only one-way protection in higher-order settings?}
Another important point is that, in a sense, local capabilities provide only
one-way protection: the caller is protected from the callee but not vice-versa.
Concretely: when invoking a function with some arguments marked as local, the
caller is guaranteed that the callee will not have been able to store the
capabilities anywhere (except perhaps on the stack, see above). However, the
callee seems to have more limited guarantees: Particularly, the caller may have
kept its own stack capability and this stack capability may (and typically will)
also cover the part of the stack that is ``owned'' by the callee.  In this
sense, the guarantees are more limited than in a linear language.

So what does this mean? In a first-order language, this is all fine, but what if
we are in a higher-order language. Imagine the following (in some ML-like language):

\begin{verbatim}
let f = fun callback =>
          let ... in 
          let ret = callback() in
          ...
//adversary top function
let advtop = f( (fun y => ...) )
\end{verbatim}

Our trusted function f is invoked by the adversary (from function advtop()) and
wants to be able to invoke an untrusted callback received from the adversary.
When invoking the closure, we don't want it to be able to access f's local
variables which it has stored on the stack. To achieve this, we only give it a
stack pointer that covers the part of the stack that is unused by f. However,
the callback may be implemented as an entry pointer that carries capabilities,
particularly the capability to advtop's stack pointer, which includes the part
of the stack that is now used by f and contains f's local variables.

So how do we deal with this? Perhaps we should use the fact that this is only
possible when f's callback argument is allocated to some part of the memory to
which advtop has store-local permissions (since the callback contains a
reference to the stack to which advtop only has a local capability). One way to
exclude the scenario is to require that callbacks are provided as non-local
capabilities. The downside of this is that local callbacks can be useful for the
caller to prevent the callee from storing them. Another way to exclude the
scenario is to require that the stack is allocated in a fixed part of the
address space and to check that callbacks point outside of this region before
invoking them.
